Bayesian non-parametric inference for multivariate peaks-over-threshold models

### 2022-06-18 results.tex
first pass at springer-nature class
statistics and computing submission

% \begin{figure}[tb]
% \begin{center}
%     \centering
%   \caption{Posterior predictive loss (Left) and energy score (right) for various models on simulated data.  Gen indicates baseline.\label{fig:simpples}}
%   \begin{subfigure}[b]{0.48\textwidth}
%     \centering
%     \includegraphics[height=2in, width=\textwidth]{./images/simulation_ppl}
%   \end{subfigure}
%   %
%   \begin{subfigure}[b]{0.48\textwidth}
%     \centering
%     \includegraphics[height=2in, width=\textwidth]{./images/simulation_es}
%   \end{subfigure}
% \end{center}
% \end{figure}

\begin{figure}[htb]
    \centering
    \caption{Energy score (top row) and Posterior Predictive Loss (bottom row) for various models on simulated data with ascending count of mixture components (indicated by plot heading) and number of dimensions (indicated by horizontal axis).  Gen indicates baseline.\label{fig:simpples}}
    \includegraphics[height=3in, width = \textwidth]{./images/simulation_score}
\end{figure}

@@

To evaluate our proposed approach for angular measure estimation 
we consider simulated datasets on $\mathbb{S}_{\infty}^{d-1}$, for 
$d$ ranging between 3 and 20. We generated each dataset from a mixture of projected Gammas, with the number of mixture components
  ranging between 3 and 12.  The generation procedure is detailed in Algorithm~\ref{algo:simulated}.
  \begin{algorithm}[ht]
    \footnotesize
    \caption{Simulated Angular Dataset Generation Routine\label{algo:simulated}}
    \For{$n_{\text{mix}}$ in $[3, 6, 9, 12]$}{
      Generate $n_{\text{mix}} \times 20$ shape Parameters $\alpha$\\
      Generate $n_{\text{mix}} \times 20$ Rate Parameters $\beta$\\
      Generate $500$ Mixture Component Identifiers $\delta$\\
      \For{$i$ in $1,\ldots,500$}{
        Generate ${\bf X}_i \sim \prod_{\ell = 1}^d\text{Ga}\left(X_{il}\mid\alpha_{\delta_i,l},\beta_{\delta_i, l}\right)$
        }
      \For{$n_{\text{col}}$ in $[3,6,12,20]$}{
        Project columns 1 to $n_{\text{col}}$ of ${\bf X}$ onto 
        $\mathcal{S}_{\infty}^{n_{\text{col}} - 1}$ and save.
        }
    }
  \end{algorithm}
Each one of the shape parameters $\alpha$ were generated 
$\alpha = \alpha_0 + \alpha_1$ where
  $\alpha_0 \sim \text{Unif}(\alpha_0\mid 0,4)$, $\alpha_1\sim \text{Gamma}(\alpha_1\mid 1,1)$.
The scale parameters $\beta$ were obtained as $\beta\sim\text{Unif}(\beta\mid 0.25, 2.5)$.

% \begin{figure}[ht]
%   \caption{Simulation Energy Score versus Column Count, by Mixture Component Count. Lower is better.\label{fig:simes}}
%   \centering
%   \includegraphics[width=0.8\textwidth]{./images/simulation_es}
% \end{figure}

@@

% We frequently see all of the gamma-based mixture models accomplish a lower energy score
%     than the generative distribution.  As \cite{nunez2019} noted, the projected Gamma model
%     is flexible and can generate multimodal distributions from a single mixture component.
%     When we allow a mixture model, we allow the possibility to break those modes into 
%     different mixture components.  This will tend to result in a better energy score, as 
%     posterior predictive replicates for a given observation in a local model will tend to 
%     be concentrated in that local mode rather than being spread between multiple modes.  
%     As for the dominance of projected restricted Gamma, By specifying $\beta_{\ell} := 1$ 
%     for all $l$, if there were multi-modal mixture components, we are in effect breaking 
%     the mixture components into separate modes. As for whether this represents 
%     \emph{overfitting}, If we had reason to believe that the generative distribution for
%     real data followed this form, then such a case could be made.  However, we are using 
%     projected Gamma as a parametric stand-in for an unknown distribution. In this sense, 
%     the argument for overfitting is less clear.
%     {\bf This paragraph is not very clear. Please make an effort to tell the story more clearly.}

@@

\begin{algorithm}[htb]
    \footnotesize
   \caption{Data preprocessing to isolate and transform data exhibiting extreme behavior.  $r_i$
   represents the radial component, and $\bm{v}_i$ the angular component.  The declustering
   portion is relevant for data correlated in time.\label{algo:processing}}
  \KwResult{$\bm{ r},\bm{v} : r_i \sim \text{Pareto}(1)$, $\bm{ v}_i \in {\mathbb S}_{\infty}^{d-1}$}
  \For{$\ell = 1,\ldots,d$}{
    Set $b_{t,\ell} = \hat{F}_{\ell}^{-1}\left(1 - \frac{1}{t}\right)$.\\
    With $\bm{ x}_{\ell} > b_{t,\ell}$, fit $a_{\ell}$, $\xi_{\ell}$ via MLE according to generalized Pareto likelihood.\\
    }
  \For{$i = 1,\ldots,n$}{
    Define $z_{i,\ell} = \left(1 + \xi_{\ell}\frac{x_{i,\ell} - b_{t,\ell}}{a_{\ell}}\right)_{+}^{1/\xi_{\ell}}$; $\;\;\;$ then $r_i = \pnorm{\bm{ z}_i}{\infty}$, $\;\;\bm{ v}_i = \frac{\bm{ z}_i}{\pnorm{\bm{ z}_i}{\infty}}$\\
    }
  Subset $\bm{ r},\bm{ v}$ such that $r_i \geq 1$\\
  \If{declustering}{
    \For{$i = 1,\ldots,n$}{
      If $r_i \geq 1$ and $r_{i-1} \geq 1$, drop the lesser (and associated $v_i$) from data set.\\
    }
  }
\end{algorithm}

\begin{figure}[ht]%{l}{0.45\textwidth}
    \centering
    \caption{Pairwise plots from ERA-Interim data after transformation and projection to ${\mathbb S}_{\infty}^{7}$.  Down the diagonal are marginal kernel densities, with two-dimensional histograms on the off-diagonal.  In those plots, red indicates a higher density.  All data is between 0 and 1.\label{fig:erai_data}}
    \includegraphics[width=.8\linewidth]{./images/data_transformed}
\end{figure}

@@

After subsetting the ERA-Interim data to the rainy season we 
  have $5587$ observations. After the processing and declustering described in
  Algorithm~\ref{algo:processing}, this number reduces to $511$ observations. A pairwise plot 
  of the transformed data after processing and declustering is presented in 
  Figure~\ref{fig:erai_data}.  From this, we note that the marginal densities display strong
  similarities, with a large spike near 0 and a small spike near 1. A value of 1 in a particular 
  axis indicates that the standardized threshold exceedance was largest in that dimension.  
  The off-diagonal plots correspond to pairwise density plots.  We observe that some site pairs, such 
  as $(1,2)$, $(7,8)$, and especially $(4,5)$ have the bulk of their data concentrated in a 
  small arc along the $45^{\circ}$, while other site combinations such as $(3,6)$, $(2,7)$, or
  $(1,8)$ the data is split, favoring one side or the other of the $45^{\circ}$ line. For the 
  ERA5 data, after subsetting we have $6342$ observations, which reduces to $532$ observations 
  after processing and declustering. We fit the PG--G, PRG--G, PG--LN, and PRG--LN models to 
  both datasets.
%  {\bf At this point it would be nice to include a figure with a plot of the data after all the pre-processing. You show a couple of panels for the ERA-Iterim and a couple of the ERA5.}
%  \makenote{Do you mean marginal histograms, or, like, a couple selected ternary scatter-plots?}

@@

\begin{figure}[ht]
    \centering
    \caption{Pairwise extremal dependence coefficients for IVT data using the PRG--G model.\label{fig:chi_ij}}
    \includegraphics[width=0.9\textwidth]{./images/chi_ij_c}
\end{figure}

@@

\begin{figure}[htb]
    \centering
    \caption{Conditional survival curves for selected locations, using ERA-Interim, and PRG-G model,  conditioning on all other dimensions at greater than 90th percentile (fitted)\label{fig:condsurv1d}. The left panel uses original units. Right panel uses standardized units.}
    \includegraphics[width=\linewidth]{./images/condsurv_1d}
\end{figure}

\begin{figure}[htb]
    \centering
    \caption{Pairwise conditional survival curves for selected locations, using ERA-Interim, and PRG-G model, conditioning on all other dimensions at greater than 90th percentile (fitted).\label{fig:condsurv2d}}
    \includegraphics[height=5in, width=5in]{./images/condsurv_2d}
\end{figure}

>>>>

\begin{figure*}[htb]
    \centering
    \caption{Energy score (top row) and Posterior Predictive Loss (bottom row) for various models on simulated data with ascending count of mixture components (indicated by plot heading) and number of dimensions (indicated by horizontal axis).  Gen indicates baseline.\label{fig:simpples}}
    \includegraphics[height=3in, width = \textwidth]{./images/simulation_score}
\end{figure*}

@@

To evaluate our proposed approach for angular measure estimation 
    we consider simulated datasets on $\mathbb{S}_{\infty}^{d-1}$, for 
    $d$ ranging between 3 and 20. We generated each dataset from a mixture of projected Gammas, with 
    the number of mixture components ranging between 3 and 12.  The generation procedure is detailed 
    in Algorithm~\ref{algo:simulated}.
    \begin{algorithm}[ht]  % algorithm using algorithm package
        \caption{Simulated Angular Dataset Generation Routine\label{algo:simulated}}
        \begin{algorithmic}
        \For{$n_{\text{mix}}$ in $[3, 6, 9, 12]$}
            \State Generate $n_{\text{mix}} \times 20$ shape Parameters $\alpha$
            \State Generate $n_{\text{mix}} \times 20$ Rate Parameters $\beta$
            \State Generate $500$ Mixture Component Identifiers $\delta$
            \For{$i$ in $1,\ldots,500$}
                \State Generate ${\bf X}_i \sim \prod_{\ell = 1}^d\text{Ga}\left(X_{il}\mid\alpha_{\delta_i,l},\beta_{\delta_i, l}\right)$
            \EndFor
            \For{$n_{\text{col}}$ in $[3,6,12,20]$}
                \State Project columns 1 to $n_{\text{col}}$ of ${\bf X}$ onto $\mathcal{S}_{\infty}^{n_{\text{col}} - 1}$ and save.
            \EndFor
        \EndFor
        \end{algorithmic}
    \end{algorithm}

Each one of the shape parameters $\alpha$ were generated 
$\alpha = \alpha_0 + \alpha_1$ where
  $\alpha_0 \sim \text{Unif}(\alpha_0\mid 0,4)$, $\alpha_1\sim \text{Gamma}(\alpha_1\mid 1,1)$.
The scale parameters $\beta$ were obtained as $\beta\sim\text{Unif}(\beta\mid 0.25, 2.5)$.

@@

@@

\begin{algorithm}[htb]
    \caption{Data preprocessing to isolate and transform data exhibiting extreme behavior.  $r_i$
    represents the radial component, and $\bm{v}_i$ the angular component.  The declustering 
    portion is relevant for data correlated in time.\label{algo:processing}}
    \begin{algorithmic}
        \For{$\ell = 1,\ldots,d$}
            \State Set $b_{t,\ell} = \hat{F}_{\ell}^{-1}\left(1 - \frac{1}{t}\right)$.
            \State With $\bm{ x}_{\ell} > b_{t,\ell}$, fit $a_{\ell}$, $\xi_{\ell}$ via MLE according to generalized Pareto likelihood.
        \EndFor
        \For{$i = 1,\ldots,n$}
            \State Define $z_{i,\ell} = \left(1 + \xi_{\ell}\frac{x_{i,\ell} - b_{t,\ell}}{a_{\ell}}\right)_{+}^{1/\xi_{\ell}}$; $\;\;\;$ then $r_i = \pnorm{\bm{ z}_i}{\infty}$, $\;\;\bm{ v}_i = \frac{\bm{ z}_i}{\pnorm{\bm{ z}_i}{\infty}}$
        \EndFor
        \State Subset $\bm{ r},\bm{ v}$ such that $r_i \geq 1$
        \If{declustering}
            \For{$i = 1,\ldots,n$}
                \State If $r_i \geq 1$ and $r_{i-1} \geq 1$, drop the lesser (and associated $v_i$) from data set.
            \EndFor
        \EndIf
    \end{algorithmic}
\end{algorithm}

\begin{figure*}[ht]%{l}{0.45\textwidth}
    \centering
    \caption{Pairwise plots from ERA-Interim data after transformation and projection to ${\mathbb S}_{\infty}^{7}$.  Down the diagonal are marginal kernel densities, with two-dimensional histograms on the off-diagonal.  In those plots, red indicates a higher density.  All data is between 0 and 1.\label{fig:erai_data}}
    \includegraphics[width=.7\linewidth]{./images/data_transformed}
\end{figure*}

@@

After subsetting the ERA-Interim data to the rainy season we 
  have $5587$ observations. After the processing and declustering described in
  Algorithm~\ref{algo:processing}, this number reduces to $511$ observations. A pairwise plot 
  of the transformed data after processing and declustering is presented in 
  Figure~\ref{fig:erai_data}.  From this, we note that the marginal densities display strong
  similarities, with a large spike near 0 and a small spike near 1. A value of 1 in a particular 
  axis indicates that the standardized threshold exceedance was largest in that dimension.  
  The off-diagonal plots correspond to pairwise density plots.  We observe that some site pairs, such 
  as $(1,2)$, $(7,8)$, and especially $(4,5)$ have the bulk of their data concentrated in a 
  small arc along the $45^{\circ}$, while other site combinations such as $(3,6)$, $(2,7)$, or
  $(1,8)$ the data is split, favoring one side or the other of the $45^{\circ}$ line. For the 
  ERA5 data, after subsetting we have $6342$ observations, which reduces to $532$ observations 
  after processing and declustering. We fit the PG--G, PRG--G, PG--LN, and PRG--LN models to 
  both datasets.

@@

\begin{figure*}[htb]
    \centering
    \caption{Pairwise extremal dependence coefficients for IVT data using the PRG--G model.\label{fig:chi_ij}}
    \includegraphics[width=0.9\textwidth]{./images/chi_ij_c}
\end{figure*}

@@

\begin{figure*}[htb]
    \centering
    \caption{Conditional survival curves for selected locations, using ERA-Interim, and PRG-G model,  conditioning on all other dimensions at greater than 90th percentile (fitted)\label{fig:condsurv1d}. The left panel uses original units. Right panel uses standardized units.}
    \includegraphics[width=\linewidth]{./images/condsurv_1d}
\end{figure*}

\begin{figure*}[htb]
    \centering
    \caption{Pairwise conditional survival curves for selected locations, using ERA-Interim, and PRG-G model, conditioning on all other dimensions at greater than 90th percentile (fitted).\label{fig:condsurv2d}}
    \includegraphics[height=4in, width=4in]{./images/condsurv_2d}
\end{figure*}
